Supporting Information : Effect of copula on extinction risk

### 2019-03-14 MainText.Rmd
wrting methods : density dep. metapop. model...continued

# Introduction

objectives:

1) Nature of tail dependence in environmental noise (or in vital rates) affects extinction risk of a population differently.
       - Age structure stochastic matrix model : extrisk of left tail dep. > extrisk of right tail dep.
         (This difference is small, could be explained by Tuljapurkar's approx. for linear model, see discussion)

       - 6 Density-dependent models with extreme tail dep. in noise (local/ global dispersal)
           - extrisk depends on tail dependnece as well as on model parameters (under-compensatory, over-compensatory)
           - Rescue effect ??
           - H0 : switching occurs near rc (without dispersal)

@@

<!--stochastic matrix model : age structured-->
To investigate how copula structure in noisy environment affect the extinction risk of
populations in a single patch age-structured model,
we considered a simple stochastic age-structured model (Eq. \ref{eq_age_str}) 
with two age classes(egg $E$ and adult $A$). We assumed
the population was semelperous, i.e., adult ($A$) went through the
reproductive episode only once before death. Eggs and adults in the 
next generation depend upon the fecundity of adult ($f_a \geq 0$) 
and survival of eggs ($0 \leq s_e \leq 1$) from present generation, respectively.
The environmental stochasticity in the model was embedded in 
the fluctuating vital rates ($f_a$ and $s_e$). We considered the fecundity ($f_a$)
and survival ($s_e$) rates were random draw from a gamma ($\Gamma(k,\theta)$) and
a beta ($\mathcal{B}(\alpha,\beta)$) distribution, respectively. These distribution 
could be generated from iid noises showing specific tail-dependence (left or right tail)
between fecundity and survival rates. Eq. \ref{eq_vitalrates} represent the stochasticity 
in the vital rates. We included the density dependence effect in one of the vital rate,
namely, fecundity ($f_a$). If $\Omega = \infty$ in Eq. \ref{eq_vitalrates}, 
the projection matrix would be independent
of populations from any age-class and the model was then called to be "linear", whereas
its "nonlinear" version had finite $\Omega$.

@@

We repeated the above analysis considering the vital rates coming from a 
an asymmetric copula with specific 
left (Clayton, Survival Gumbel)
or right (Survival Clayton, Gumbel) tail dependence
or from an extreme tail-depentdent one, 
when vital rates were considered to be affected
by extreme climatic events. A $N \times 2$ matrix, M, was generated
using `BiCopSim` function
from `VineCopula` package 
for each of the left 
(Clayton, Survival Gumbel) and right 
(Survival Clayton, Gumbel) tail-dependent bivariate copulas, 
where $N = numsteps(=25) \times numsims(=10000)$.
We kept the same Kendall's correlation ($\tau = 0.6$) for 
each bivariate copula from one parameter family which used its parameter accordingly
(see `iTau` function from `copula` package). Entries within each clumn of
matrix $M$, were thus 
independent of each others but 
the two columns of $M$ had a specified Kendall's 
correlation ($\tau = 0.6$) and underlying tail-dependence. 
Similarly, for extreme-tail dependence
(either left or right tail) a matrix M with two column was generated as described in 
Appendix \ref{SM-extr_tail_dep}.

@@

<!--density-dependence metapopulation model-->

<!--calculation of ext risk with density-dependence metapopulation model-->

@@

<!--FIG :  summary : all nonlinear model results for local dispersion : extrisk vs. d-->
\begin{figure*}[!h]
\begin{center}
\includegraphics[width=15 cm]{./Results/allmodel_ext_risk_local_disp.pdf}
\caption{For each of six nonlinear density-dependence models presented in Table \ref{tab_model_summary} with local dispersal, extinction risk varies as dispersion-rate ($d$) goes from $0$ to $0.5$. Last two columns represent extinction risk calculated for bifurcation parameter $(r)$ taken from under-compensatory and over-compensatory regime of deterministic model analog, respectively.\label{fig_extrisk_allmodel_localdisp}}
\end{center}
\end{figure*}

@@

<!--Summary tables for 6 nonlinear models used-->
```{r read_tab_model_summary,echo=F, results='asis',message=F}
library(kableExtra)
library(dplyr)
dt<-readRDS("./Results/tab_model_summary.RDS")
knitr::kable(dt, escape=F,
             format="latex",digits=3, align="c",table.env='table*',
             caption = "Summary of six density dependence models considered for each metapopulation patch with local or global dispersal; last three columns indicated equilibrium population value ($P_e$), critical growth rate ($r_c$) where deterministic analog of models showed transition from under- to over-compensatory dynamics and bifurcation parameter ($r_{bf}$) where $P_e$ became unstable from stable steady state, respectively.\\label{tab_model_summary}",
             booktabs=T, linesep = "\\addlinespace"
             )%>%column_spec(c(1,3,5,6),width="1.4cm")%>%kable_styling(latex_options = "basic",position = "center")%>%column_spec(2, width = "3.5 cm")
```

>>>>

# Introduction\label{intro}

<!--
Nature of tail dependence in environmental noise (or in vital rates) affects extinction risk of a population differently.
       - Age structure stochastic matrix model : extrisk of left tail dep. > extrisk of right tail dep.
         (This difference is small, could be explained by Tuljapurkar's approx. for linear model, see discussion)

       - 6 Density-dependent models with extreme tail dep. in noise (local/ global dispersal)
           - extrisk depends on tail dependnece as well as on model parameters (under-compensatory, over-compensatory)
           - Rescue effect ??
           - H0 : switching occurs near rc (without dispersal)
-->

We explored the potential of asymmetric tail-dependence in copula 
for understanding the extinction risk of populations towards extreme climatic event in ecology
by addressing the following specific questions: (Q1) Does nature of tail dependence in vital rates
of a single-patch age-structured population model affect extinction risk of that population differently?
We hypothesized population might have higher extinction risk if vital rates are left-tail dependent.
(Q2) How does asymmetric tail-dependent noise across multiple patches influence extinction 
risk of a nonlinear metapopulation model having density dependence? We hypothesized the general finding 
from density-independent metapopulation model (higher extinction risk if noise across patches were 
left-tail dependent than the one if noise came from right-tail dependent) holds true only in
the under-compensatory regime, but should be reverse if the model shows over-compensatory dynamics. 
(Q3) We further hypothesize that when there is no dispersal, the above-said switching should occur
near the critical parameter value where the deterministic model shows transition from 
under-compensatory to over-compensatory regime.

@@

<!--stochastic matrix model : age structured-->
To investigate (Q1) from \nameref{intro} that how copula structure in noisy 
environment affect the extinction risk of
populations in a single patch age-structured model,
we considered a simple stochastic age-structured model (Eq. \ref{eq_age_str}) 
with two age classes (egg $E$ and adult $A$). We assumed
the population was semelperous, i.e., adult ($A$) went through the
reproductive episode only once before death. Eggs and adults in the 
next generation depend upon the fecundity of adult ($f_a \geq 0$) 
and survival of eggs ($0 \leq s_e \leq 1$) from present generation, respectively.
The environmental stochasticity in the model was embedded in 
the fluctuating vital rates ($f_a$ and $s_e$). We considered the fecundity ($f_a$)
and survival ($s_e$) rates were random draw from a gamma ($\Gamma(k,\theta)$) and
a beta ($\mathcal{B}(\alpha,\beta)$) distribution, respectively. These distribution 
could be generated from iid noises showing specific tail-dependence (left or right tail)
between fecundity and survival rates. Eq. \ref{eq_vitalrates} represent the stochasticity 
in the vital rates. We included the density dependence effect in one of the vital rate,
namely, fecundity ($f_a$). If $\Omega = \infty$ in Eq. \ref{eq_vitalrates}, 
the projection matrix would be independent
of populations from any age-class and the model was then called to be "linear", whereas
its "nonlinear" version had finite $\Omega$.

@@

To test our hypothesis from (Q1), we repeated the above analysis 
considering the vital rates coming from a 
an asymmetric copula with specific 
left (Clayton, Survival Gumbel)
or right (Survival Clayton, Gumbel) tail dependence
or from an extreme tail-depentdent one, 
when vital rates were considered to be affected
by extreme climatic events. A $N \times 2$ matrix, M, was generated
using `BiCopSim` function
from `VineCopula` package 
for each of the left 
(Clayton, Survival Gumbel) and right 
(Survival Clayton, Gumbel) tail-dependent bivariate copulas, 
where $N = numsteps(=25) \times numsims(=10000)$.
We kept the same Kendall's correlation ($\tau = 0.6$) for 
each bivariate copula from one parameter family which used its parameter accordingly
(see `iTau` function from `copula` package). Entries within each clumn of
matrix $M$, were thus 
independent of each others but 
the two columns of $M$ had a specified Kendall's 
correlation ($\tau = 0.6$) and underlying tail-dependence. 
Similarly, for extreme-tail dependence
(either left or right tail) a matrix M with two column was generated as described in 
Appendix \ref{SM-extr_tail_dep}.

@@

<!--density-dependence metapopulation model-->
We explored (Q2) by examining how nature of the tail dependence of an
environmental variable across space could influence the extinction risk of a 
metapopulation model with density-dependence feedback. We considered six density-dependent 
population-growth models (Ricker, Hassell, Maynard Smith, Pennycuick, Verhulst, Austin-Brewer)
in noisy environment (see Table \ref{tab_model_summary}) 
from a metapopulation context.
Models were adapted from earlier work
by @cohen1995unexpected, excluding only two : Varley model (this is not a continuous one) 
and Malthus-Condorcet-Mill model (terms containing logarithm of extinct population does not make sense). 
All the models could be represented via 
a generalized form : $\vec{P_{t+1}} = D \lambda_t \vec{P_t}$,
where $\vec{P_t}$ was a vector of population densities, 
the $i^{th}$ component $P_t^i$ representing
the density in the $i^{th}$ habitat patch ($i=1,\ldots,n$) at time $t$. 
The $n \times n$ matrix $\lambda(t)$ was diagonal with $i^{th}$ diagonal entry 
of stochastic density dependence form $\lambda_i(t)$ (see second column 
of Table \ref{tab_model_summary}). For simplicity, 
we considered $n$ patches in each model were arranged, evenly spaced, in a line
with the provision of "local" or "global" dispersal among patches. A n $\times$ n dispersal
matrix, $D$ was constructed assuming that a fraction $d$ of 
each population patch dispersed during any given time step, equally distributed to
the two or one nearest neighbors in case of local dispersal. For global dispersal, the fraction
$d$ that dispersed
from each patch was divided up equally among all other patches. We called $d$ as dispersal
rate with $d$ being zero there was no dispersal, and threby no chance to rescue an extinct 
patch through migration of nearby metapopulation patches.

The environmental noises from any two patch, say, 
$(\epsilon_t^i, \epsilon_t^j)$ were normally 
distributed within each patch with mean $m=0$ and standard deviation $sd$, 
(see the specified $sd$ for each model in 3$^{rd}$ 
column of Table \ref{tab_model_summary}), were 
independent through time, and showed the same 
spatial correlations between patches (to within sampling error) for every simulation. 
But they were made to exhibit right- or left-tail dependence between patches in
different simulations. A three dimensional array, numsims $\times$ n $\times$ numsteps, 
$\epsilon$, was made from these extreme tail-dependence noise 
matrix generated via algorithm stated in 
Appendix \ref{SM-extr_tail_dep} and used to calculate extinction risk 
for each of the specified model in Table \ref{tab_model_summary} for $numsteps = 25$ years, $n =$ 5 patches
and $numsims = 10000$ simulations.

<!--say how you set ext. threshold = Pe/10-->

<!--test hypothesis from Q2 : ext risk with density-dependence metapopulation model with r from different regime-->

<!--test hypothesis from Q3 : switching of ext risk with density-dependence metapopulation model with d=0-->

@@

<!--FIG :  summary : all nonlinear model results for local dispersion : extrisk vs. d-->
\begin{figure*}[!h]
\begin{center}
\includegraphics[width=15 cm]{./Results/allmodel_ext_risk_local_disp.pdf}
\caption{For each of six nonlinear density-dependence models presented in Table \ref{tab_model_summary} with local dispersal
among $n=5$ patches, extinction risk after $numsteps = 25$ years for $numsims = 10000$ simulations were calculated.
Extinction risk varies as dispersion-rate ($d$) goes from $0$ to $0.5$. Last two columns represent extinction risk calculated for intrinsic growth rate $(r)$ taken from under-compensatory and over-compensatory regime of deterministic model analog, respectively.\label{fig_extrisk_allmodel_localdisp}}
\end{center}
\end{figure*}

@@

<!--Summary tables for 6 nonlinear models used-->
```{r read_tab_model_summary,echo=F, results='asis',message=F}
library(kableExtra)
library(dplyr)
dt<-readRDS("./Results/tab_model_summary.RDS")
knitr::kable(dt, escape=F,
             format="latex",digits=3, align="c",table.env='table*',
             caption = "Summary of six density dependence metapopulation models considered; last three columns indicated equilibrium population value ($P_e^{i}$), critical growth rate ($r_c^{i}$) where deterministic analog of models showed transition from under- to over-compensatory dynamics and bifurcation parameter ($r_{bf}^{i}$) where $P_e^{i}$ became unstable from stable steady state, respectively. Superscript $i=1,2,...,n$ indicates states for $i^{th}$ patch.\\label{tab_model_summary}",
             booktabs=T, linesep = "\\addlinespace"
             )%>%column_spec(c(1,3,5,6),width="1.5 cm")%>%kable_styling(latex_options = "basic",position = "center")%>%column_spec(2, width = "3.8 cm")
```

\clearpage

# Literature cited

### 2019-03-14 SuppMat.Rmd
wrting methods : density dep. metapop. model...continued

dt1 <- tibble(Models = c("Ricker",
           "Hassell",
           "Maynard Smith",
           "Pennycuick",
           "Verhulst",
           "Austin-Brewer")
)

dt2<-tibble(Equations = c("$P_{t+1} = \\exp(\\epsilon_t) \\hspace{0.1 cm} P_t \\exp[r (1-P_t/K)$]",
                          "$P_{t+1} = \\exp(\\epsilon_t) \\hspace{0.1 cm} rP_t/(1+aP_t)^b$",
                          "$P_{t+1} = \\exp(\\epsilon_t) \\hspace{0.1 cm} rP_t/[1+(aP_t)^b]$",
                          "$P_{t+1} = \\exp(\\epsilon_t) \\hspace{0.1 cm} rP_t/[1+\\exp \\{-a(1-P_t/b)\\}]$",
                          "$P_{t+1} = \\exp(\\epsilon_t) \\hspace{0.1 cm} P_t [1+r(1-P_t)]$",
                          "$P_{t+1} = \\exp(\\epsilon_t) \\hspace{0.1 cm} P_t [1+r(K-P_t)\\{1-\\exp(-sP_t)\\}]$"
))

@@

dt4<-tibble("$P_e$" = c("K",
                        "$(r^{1/b}-1)/a$",
                        "$[(r-1)^{1/b}]/a$",
                        "$[b\\{a+ln(r-1)\\}]/a$",
                        "1",
                        "K"
))

dt5<-tibble("$r_c$" = c(rc_ricker,
                        rc_hassell,
                        rc_msmith,
                        rc_pennycuick,
                        rc_verhulst,
                        rc_abrewer
))

dt6<-tibble("$r_{bf}$" = c(r_bf_ricker,
                           r_bf_hassell,
                           r_bf_msmith,
                           r_bf_pennycuick,
                           r_bf_verhulst,
                           r_bf_abrewer
))

@@

>>>>

dt1 <- tibble("Models $(P_{t+1}^{i} = \\lambda_t^{i} P_t^{i})$"= c("Ricker",
           "Hassell",
           "Maynard Smith",
           "Pennycuick",
           "Verhulst",
           "Austin-Brewer")
)

dt2<-tibble("Density dependence $(\\lambda_t^{i})$ with noise" = c("$\\exp(\\epsilon_t^{i}) \\hspace{0.1 cm} \\exp[r (1-P_t^{i}/K)$]",
                          "$\\exp(\\epsilon_t^{i}) \\hspace{0.1 cm} r/(1+aP_t^{i})^b$",
                          "$\\exp(\\epsilon_t^{i}) \\hspace{0.1 cm} r/[1+(aP_t^{i})^b]$",
                          "$\\exp(\\epsilon_t^{i}) \\hspace{0.1 cm} r/[1+\\exp \\{-a(1-P_t^{i}/b)\\}]$",
                          "$\\exp(\\epsilon_t^{i}) \\hspace{0.1 cm} [1+r(1-P_t^{i})]$",
                          "$\\exp(\\epsilon_t^{i}) \\hspace{0.1 cm} [1+r(K-P_t^{i})\\{1-\\exp(-sP_t^{i})\\}]$"
))

@@

dt4<-tibble("$P_e^{i}$" = c("K",
                        "$(r^{1/b}-1)/a$",
                        "$[(r-1)^{1/b}]/a$",
                        "$[b\\{a+ln(r-1)\\}]/a$",
                        "1",
                        "K"
))

dt5<-tibble("$r_c^{i}$" = c(rc_ricker,
                        rc_hassell,
                        rc_msmith,
                        rc_pennycuick,
                        rc_verhulst,
                        rc_abrewer
))

dt6<-tibble("$r_{bf}^{i}$" = c(r_bf_ricker,
                           r_bf_hassell,
                           r_bf_msmith,
                           r_bf_pennycuick,
                           r_bf_verhulst,
                           r_bf_abrewer
))

@@

# Stability analysis of deterministic metapopulation models  \label{model_stability}
